The Structure of Human Language Reflects Limitations of Working Memory

### 2019-05-15 paper-short.tex

\usepackage{showlabels}
\usepackage{fullpage}
\usepackage{pslatex}
%\usepackage{latexsym}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{url}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{rotating}
\usepackage{amssymb}

@@

\newcommand{\thetad}[0]{{\theta_d}}
\newcommand{\thetal}[0]{{\theta_{LM}}}

@@

\begin{abstract}
  %  the Nature template:
  % 1-2 sentences basic introduction, for anyone
  Each of the roughly 6000 natural languages currently spoken in the world embodies a different system for encoding thoughts into linear strings of words.
  By characterizing the commonalities and differences among these systems we can gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence. 
  % 2-3 sentences detailed background, for scientists in related disciplines (linguistics, cogsci, computer science, info theory)
  In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension \cite{gibson2019efficiency,florian,hawkins,piantadosi2011word,zipf1949}.

  % 1 sentence general problem
  % 1 sentence here we show
  Here we give evidence for this view by showing that human languages structure information in time so as to enable efficient communication under the constraint that language must be comprehended and produced by agents with limited short-term memory resources.
  We directly show that languages minimize memory usage by calculating information-theoretic bounds on the amount of memory resources required for language comprehension in 54 real languages and in random baseline languages, finding that the real languages require less memory.
  % 2-3 sentences how main result adds to knowledge
  To explain how languages achieve this result, we prove a theorem showing that any efficient communication system operating under constraints on memory capacity will exhibit a property called information locality: symbols which predict each other should be produced close to each other in time.
  We show that information locality subsumes several principles of word order previously proposed in the linguistics and cognitive science literature, and explains a number of the previously-documented universal patterns in word order across languages.
	% TODO i'm wondering whether we should first mention the big result, and then say, well it also subsumes previously proposed constraints.
  % 1-2 sentences general context

@@

% introductory paragraph

\note{all three of the next are like first sentences}
Human languages express thoughts into sentences, i.e. linear strings of words. 
%For example, if your mountain climber friend Bob decided to turn back near the summit of Mt. Everest, you might say \emph{Bob gave the attempt to climb Mt. Everest up}.
% Run with this example
\jd{is the commented out example sth that will make it in? it does seem like an example would be good here}

Each language embodies a different strategy for encoding thoughts into strings of words. \jd{will this sentence and the next both survive?}\note{make this sentence nonredundant}
\note{give an example}

@@

@@

We begin developing our model by considering the process of language comprehension in a person processing a stream of words uttered by an interlocutor (Figure~\ref{fig:communication}).
Experimental research has established that such a listener (1) maintains information about the words received so far, (2) forms probabilistic expectations about the upcoming words, and (3) words are easier to process when they are highly expected in context \citep{hale2001probabilistic,levy2008expectation} \rljf{Good source for other cites here would be to dig around in Luke \& Christianson's review of predictability effects}.

Experimental research has shown that human processing effort on a given word is strongly predicted by its \textbf{surprisal}, the degree to which it cannot be predicted from the preceding context \citep{hale2001probabilistic,levy2008expectation}.
Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times \citep{smith2013effect}.
For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
\begin{equation}\label{eq:surp}
	S :=	-\log_2 p(x_i|x_{1...i-1})
\end{equation}
where $x_i$ is the $i$-th word in a given utterance (or discourse). In information-theoretic terms, surprisal is equivalent to the \textbf{information content} of a word, measured in bits, and the fact that the per-word reading time appears to be linear in the surprisal suggests that human language comprehension is best characterized in terms of a scalar rate of information processing in terms of bits per second. The idea that processing effort is determined by surprisal links language comprehension to theories of perception and brain function that are organized around the idea of prediction and \key{predictive coding} \citep{friston2009predictive,friston2010free,clark2013whatever}, which has had empirical success in other domains such as perception \citep{egner2010expectation}, auditory and music perception \citep{agres2018information}, and motor planning \citep{wolpert2001motor}.

However, Equation~\ref{eq:surp} can only represent processing difficulty for a listener with unbounded memory resources, who can accurately remember the complete history of previous words $x_{1...i-1}$ leading up to the current word $x_i$. More realistically, human listeners have fixed-capacity memory resources that are being deployed during language comprehension, meaning that their predictions about upcoming words may be inaccurate compared to Eq.~\ref{eq:surp}. If $m_{i-1}$ is a listener's memory state before processing word $x_i$, then the true surprisal experienced by the listener will be given by \ref{eq:lossy-surp}:
\begin{equation}
  \label{eq:lossy-surp}
  S_M := -\log_2 p(x_i|m_{i-1}),
\end{equation}
which will on average be larger than Eq.~\ref{eq:surp} (see SI for proof). % Gibbs' inequality
The limitations of memory have well-documented empirical effects on language processing, causing difficulty above and beyond the difficulty associated with surprisal \citep{gibson1998linguistic,gibson1999memory,gibson2000dependency,vasishth2005activationbased,levy2013memory}.

@@

\rljf{I am feeling the urge to reframe everything in terms similar to my NYU talk, where we say the processing effort of a language is its entropy rate (an unavoidable cost) plus excess cost arising from memory distortion, which should be minimized. This feels like a very general theory, which can be explained graphically. Linguists and psycholinguists will not understand it at first, but maybe they will if we include the stuff at the end about how this relates to dependency locality in word order and processing. The entropy rate framing would allow us to define more useful terms for explaining our results. For example above we have $x_i$, $x_{1...i-1}$, and $m_{i-1}$ as values, but really they are best thought of as random variables, becuase all of our results are about entropy and mutual information, not pointwise surprisal or pointwise MI. Maybe we could even make some kind of call-out to the competence/performance distinction: entropy rate is the best achievable efficiency in language processing given a perfect listener (like competence), but the memory-distorted cross-entropy rate gives us the true difficulty (performance). Tagline: Human languages minimize the divergence between competence and performance given fixed processing resources.}

>>>>

\usepackage{showlabels}
\usepackage{fullpage}
\usepackage{pslatex}
%\usepackage{latexsym}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{natbib}
\usepackage{url}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{rotating}
\usepackage{amssymb}

@@

\newcommand{\thetad}[0]{{\theta_d}}
\newcommand{\thetal}[0]{{\theta_{LM}}}
\newcommand{\key}[1]{\textbf{#1}}

@@

\begin{abstract}
  %  the Nature template:
  % 1-2 sentences basic introduction, for anyone
  Each of the roughly 6000 natural languages currently spoken in the world embodies a different system for encoding thoughts into linear strings of words.
  By characterizing the commonalities and differences among these systems we can gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence. 
  % 2-3 sentences detailed background, for scientists in related disciplines (linguistics, cogsci, computer science, info theory)
  In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension.

  % 1 sentence general problem
  % 1 sentence here we show
  Here we give a novel formalization of this view in terms of information theory, and demonstrate that human languages structure information in time so as to enable efficient communication under the constraint that language must be comprehended and produced by agents with limited short-term memory resources.
  We directly show that languages minimize memory usage by calculating information-theoretic bounds on the amount of working memory resources required for language comprehension in 54 real languages and in random baseline languages, finding that the real languages require less memory.
  % 2-3 sentences how main result adds to knowledge
  To explain how languages achieve this result, we prove a theorem showing that any efficient communication system operating under constraints on memory capacity will exhibit a property called information locality: symbols which predict each other should be produced close to each other in time.
  We show that information locality subsumes several principles of word order previously proposed in the linguistics and cognitive science literature, and explains a number of the previously-documented universal patterns in word order across languages.
	% TODO i'm wondering whether we should first mention the big result, and then say, well it also subsumes previously proposed constraints.
  % 1-2 sentences general context

@@

% introductory paragraph
Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.

The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \citep{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.

@@

% Nice phrases from Sims (2018):
% biological information processing should seek to maximize performance subject to constraints on information processing capacity.
% The current results can be concisely stated as follows:
% Rate-distortion theory provides analytical and numerical tools for solving such constrained optimization problems (12–14).

@@

We begin developing our model by considering the process of language comprehension illustrated in Figure~\ref{fig:communication}, where the listener Alice is processing a stream of words uttered by an interlocutor Bob (Figure~\ref{fig:communication}).
Experimental research has established that (1) listeners maintain information about the words received so far, (2) listeners forms probabilistic expectations about the upcoming words, and (3) words are easy to process to the extent that they are predictable in context \rljf{Good source for other cites here would be to dig around in Luke \& Christianson's review of predictability effects}. Furthermore, the exact mathematical form of the relationship between predictability and processing difficulty (as measured e.g. in reading times) seems to be given by the \key{information content} of a word in context \citep{hale2001probabilistic,levy2008expectation,smith2013effect}.
For an idealized listener with unbounded memory resources, information content will be given by the negative log-probability of a word given its preceding context:
\begin{equation}\label{eq:surp}
	S :=	-\log_2 p(x_i|x_{1...i-1})
\end{equation}
where $x_i$ is the $i$-th word in a given utterance (or discourse). Information content is measured in bits and gives the amount of information communicated by the word. The idea that processing effort is determined by surprisal links language comprehension to theories of perception and brain function that are organized around the idea of prediction and \key{predictive coding} \citep{friston2009predictive,friston2010free,clark2013whatever}, which has had empirical success in other domains such as perception \citep{egner2010expectation}, auditory and music perception \citep{agres2018information}, and motor planning \citep{wolpert2001motor}. The average information-processing difficulty associated with a \emph{language} would then be the expected information content of each word in context, which is equivalent to an information-theoretic quantity called \key{entropy rate}: \rljf{the def. below is true only for a stationary process but maybe that's too much detail}
\begin{equation}\label{eq:entropy-rate}
  h := \mathbb{E}[-\log_2 p(x_i|x_{1...i-1})].
\end{equation}
The entropy rate of a language (seen as a stochastic process) is thus a measurement of a component of the information-processing cost associated with comprehending the language.

However, Equation~\ref{eq:surp} can only represent processing difficulty for a listener with unbounded memory resources, who can accurately remember the complete history of previous words $x_{1...i-1}$ leading up to the current word $x_i$. More realistically, human listeners have fixed-capacity memory resources that are being deployed during language comprehension, meaning that their predictions about upcoming words will be inaccurate compared to Eq.~\ref{eq:surp}. If $m_{i-1}$ is a listener's memory state before processing word $x_i$, then the true surprisal experienced by the listener will be given by \ref{eq:lossy-surp}:
\begin{equation}
  \label{eq:lossy-surp}
  S_M := -\log_2 p(x_i|m_{i-1}),
\end{equation}
which will on average be larger than Eq.~\ref{eq:surp} (see SI for proof). The limitations of memory have well-documented empirical effects on language processing, causing difficulty above and beyond the difficulty associated with raw information content \citep{gibson1998linguistic,gibson1999memory,gibson2000dependency,vasishth2005activationbased,levy2013memory}.
The average processing difficulty associated with a language then comes out to a \key{cross entropy}:

@@

\rljf{I am feeling the urge to reframe everything in terms similar to my NYU talk, where we say the processing effort of a language is its entropy rate (an unavoidable cost) plus excess cost arising from memory distortion, which should be minimized. This feels like a very general theory, which can be explained graphically. Linguists and psycholinguists will not understand it at first, but maybe they will if we include the stuff about how this relates to dependency locality in word order and processing. The entropy rate framing would allow us to define more useful terms for explaining our results. For example above we have $x_i$, $x_{1...i-1}$, and $m_{i-1}$ as values, but really they are best thought of as random variables, becuase all of our results are about entropy and mutual information, not pointwise surprisal or pointwise MI. Maybe we could even make some kind of call-out to the competence/performance distinction: entropy rate is the best achievable efficiency in language processing given a perfect listener (like competence), but the memory-distorted cross-entropy rate gives us the true difficulty (performance). Tagline: Human languages minimize the divergence between competence and performance given fixed processing resources.}

\rljf{Here we recast these theories in terms of information theory and the theory of statistical complexity...}
